---
title: "Lab #4: Visualization Aesthetics (30 pts)"
author: "Eliel Toukam Nkamta"
format:
  html:
    code-fold: false
execute: 
  echo: true
knitr:
  opts_chunk: 
    message: false
editor: source
---

## Introduction

Use the code below to load and wrangle the data:

```{r}
#| warning: false
#| document: show
library(tidyverse)
library(RColorBrewer)

hmda_ms_2024 <- read.csv("https://ffiec.cfpb.gov/v2/data-browser-api/view/csv?states=MS&years=2024") |>
  filter(loan_type == 1 & 
           lien_status == 1 & 
           conforming_loan_limit == 'C' & 
           derived_dwelling_category %in% c("Single Family (1-4 Units):Site-Built", 
                                            "Single Family (1-4 Units):Manufactured") & 
           business_or_commercial_purpose == 2 &
           loan_purpose %in% c(1, 2)) |>
  select(lei,
         county_code,
         derived_ethnicity,
         derived_race,
         derived_sex,
         action_taken,
         loan_purpose,
         loan_amount,
         interest_rate,
         property_value,
         loan_to_value_ratio,
         denial_reason.1,
         denial_reason.2,
         denial_reason.3,
         denial_reason.4,
         income,
         debt_to_income_ratio) |>
  mutate(loan_purpose = recode(loan_purpose, 
                               "1" = "Home Purchase",
                               "2" = "Home improvement",
                               .default = NA_character_)) |>
  mutate(action_taken = recode(action_taken, 
                               "1" = "Loan originated", 
                               "2" = "Application approved but not accepted",
                               "3" = "Application denied",
                               "4" = "Application withdrawn by applicant",
                               "5" = "File closed for incompleteness",
                               "6" = "Purchased loan",
                               "7" = "Preapproval request denied",
                               "8" = "Preapproval request approved but not accepted",
                               .default = NA_character_)) |>
  mutate(denial_reason.1 = recode(denial_reason.1, 
                                  "1" = "Debt-to-income ratio",
                                  "2" = "Employment history",
                                  "3" = "Credit history",
                                  "4" = "Collateral",
                                  "5" = "Insufficient cash (downpayment, closing costs)",
                                  "6" = "Unverifiable information",
                                  "7" = "Credit application incomplete",
                                  "8" = "Mortgage insurance denied",
                                  "9" = "Other",
                                  "10" = "Not applicable",
                               .default = NA_character_)) |>
  rowwise() |>
  mutate(denial_reason = case_when(any(!is.na(c_across(c(denial_reason.2,
                                                         denial_reason.3,
                                                         denial_reason.4)))) ~ "Multiple",
      TRUE ~ denial_reason.1)) |>
  ungroup() |>
  mutate(loan_to_value_ratio = as.numeric(loan_to_value_ratio),
         property_value = as.numeric(property_value)) |>
  mutate(debt_to_income_ratio = case_when(debt_to_income_ratio %in% c("36", "37", "38","39","40","41") ~ "36%-41%",
                                          debt_to_income_ratio %in% c("42","43","44","45","46","47","48","49") ~ "42%-49%",
                                          TRUE ~ debt_to_income_ratio)) |>
  mutate(debt_to_income_ratio = factor(debt_to_income_ratio, levels = c("Exempt",
                                                                        "<20%",
                                                                        "20%-<30%",
                                                                        "30%-<36%",
                                                                        "36%-41%",
                                                                        "42%-49%",
                                                                        "50%-60%",
                                                                        ">60%"))) |>
  select(-c(denial_reason.1, 
            denial_reason.2, 
            denial_reason.3, 
            denial_reason.4))

```


::: {.callout-important}
###### Question 1

In the code below, create a bar plot that visualizes how many home loans were for home purchases vs. the number for home improvements. Determine the variable that should appear on the x-axis, along with the geom function that can be used to create a bar plot. Fill in the blanks to ensure your plot has the appropriate context.

```{r}
#| eval: false
hmda_ms_2024 |>
  ggplot(aes(x = loan_purpose)) +
  geom_bar() +
  labs(title = "Home Mortgage Disclosure Act Applications, MIssissippi, 2024",
       x = "[Loan Purpose]",
       y = "Count of Loans") +
  theme_minimal()
```
:::

::: {.callout-important}
###### Question 2

Create a plot below that counts the number of applicants of each race. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. You can use the code you wrote above as a template to get you started. You'll probably notice that the labels of your plot overlap. You can add `+ coord_flip()` to the end of your code to flip the plot so that the bars are horizontal rather than vertical. This should help with legibility.

```{r}
#| eval: false
hmda_ms_2024 |>
  ggplot(aes(x = derived_race)) +
  geom_bar() +
  labs(title = "Racial Description of Loan Seekers in the Home Mortgage Disclosure Act Applications, MIssissippi, 2024, ",
       x = "Race",
       y = "Count of Race") +
  theme_minimal()+
  coord_flip()

```

:::

::: {.callout-important}
###### Question 3

In the code below, create a histogram that visualizes the distribution of loan amounts. Determine the variable that should appear on the x-axis, along with the geom function that can be used to create a histogram. Set the binwidth to create numeric intervals of 100000. Fill in the blanks to ensure your plot has the appropriate context.

```{r}
#| eval: false
hmda_ms_2024  |> 
  ggplot(aes(x = loan_amount)) +
  geom_histogram(binwidth = 100000) +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Loan Amount",
       y = "Count of Loans",
       caption = str_wrap("This histogram displays the distribution of home loan amounts requested by Mississippi applicants in 2024. The plot shows that the majority of applicants requested loans under $250,000. Since the average home loan size in the U.S. in 2024 was $252,505, this indicates that average requested home loan sizes in MS were smaller in comparison to the country as a whole.", 120)) +
  theme_minimal()
```

:::

::: {.callout-important}
###### Question 4

Create a plot below that shows the distribution of loan to value ratios (in intervals of 10). Be sure to give it a descriptive title and labels covering all 5 essential components of data context. Also add a caption to describe the percentage of a loan's property that borrowers most often request when applying for a loan.

You can use the code you wrote above as a template to get you started. Note that there were 20 outlier loans where borrowers requested more than double property's value. There was even one loan where the applicant requested 14 times the property's value. All of these loans were denied. Below I've given you some code to filter out these outliers in order to make the plot more legible. 

```{r}
#| eval: false

outlier_filtered <- hmda_ms_2024 |>
  filter(loan_to_value_ratio < 200) 
```


```{r}
ggplot(data= outlier_filtered,
        aes(x = loan_to_value_ratio))+
  geom_histogram(binwidth = 10) +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Loan to Value Ratio",
       y = "Count")+
  theme_minimal()
  
```

:::

::: {.callout-important}
###### Question 5

Create a stacked barplot below that counts both the race of the applicants and the action taken in regards to the loan. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. I've given you some code below to move the legend to the bottom of the plot, to reduce the legend size, and to set the number of rows that appear in the legend. 

```{r}
#| eval: false

hmda_ms_2024 |>
ggplot(aes(x = derived_race,
           fill= action_taken)) +
  geom_bar() +
  labs(title = "Racial Description and Loan Action in the Home Mortgage Disclosure Act Applications, MIssissippi, 2024, ",
       x = "Race",
       y = "Count of Race") +
  theme_minimal()+
  coord_flip() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm")) +
  guides(fill=guide_legend(nrow=4))
```

:::

::: {.callout-important}
###### Question 6

Copy the plot that you created above into the code chunk below, and then adjust the *position* of the bars to "fill" in order to normalize their height. Add a caption that summarizes your findings. 

```{r}
#| eval: false

hmda_ms_2024 |>
ggplot(aes(x = derived_race,
           fill= action_taken)) +
  geom_bar(position = "fill") +
  labs(title = "Racial Description and Loan Action in the Home Mortgage Disclosure Act Applications, MIssissippi, 2024",
       x = "Race",
       y = "Count of Race",
       caption = str_wrap ("This bar plot describes how racial identity was considered by a bank in MIssissippi to grant loans. It shows that the highest number of loans were granted to the Asians followed by whites where as the loan application was denied to mainly minority races followed by Black or African American"))+
  theme_minimal()+
  coord_flip() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm")) +
  guides(fill=guide_legend(nrow=4))
  

```

:::


::: {.callout-important}
###### Question 7

Fill in the code below to create a boxplot that visualizes how the distributions of requested loan amounts differed across race. Add a caption that summarizes your findings.  

```{r}
#| eval: false

hmda_ms_2024 |>
  ggplot(aes(x = derived_race, y= loan_amount)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Variation of Race with Loan Amount from Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Loan Amount",
       caption = ("This boxplot visualizes the variation between loan amounts and different races. It shows that averagely most asians demanded for loans close to 250,000$ and most black or African Americans demanded loans less than 125,000$")) +
  theme_minimal() 
  
```

:::

::: {.callout-important}
###### Question 8
```{r}
hmda_denial_by_race <-
  hmda_ms_2024 |>
  group_by(derived_race) |>
  summarize(total_denied = sum(action_taken == "Application denied"),
            percent_denied = sum(action_taken == "Application denied")/n())

hmda_denial_by_race_sex <-
  hmda_ms_2024 |>
  filter(derived_race != "Free Form Text Only") |> #There were only 3 denials with free form text for race in MS in 2024, and they have been removed here for plot legibility
  group_by(derived_race, derived_sex) |>
  summarize(total_denied = sum(action_taken == "Application denied"),
            percent_denied = sum(action_taken == "Application denied")/n())

hmda_denial_by_race
```

In the code below, fill in the blanks to set the height of each bar to the percent of individuals in a particular race whose mortgage was denied. Select a geom function to create a column chart. Be sure to fill in the correct labels.

```{r}
#| eval: false
hmda_denial_by_race |>
  ggplot(aes(x = derived_race, y= percent_denied)) +
  geom_col()+
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Percent of Individuals Denied In Race",
       caption = str_wrap("This column plot visualizes the denial rates of home mortgage applications in MS in 2024. The plot shows that Black or African American applicants, American Indian or Alaska Native applicants, or applicants of 2 or more minority races were all denied mortgage at double the rate of White applicants. This may indicate disparities in lending practices.", 120)) +
  theme_minimal() +
  coord_flip()
  
```

:::


::: {.callout-important}
###### Question 9

Copy your code above, replacing the `hmda_denial_by_race` data frame with `hmda_denial_by_race_sex`. Using the `fill` aesthetic, add the variable for describing the applicants' sex to the plot. Set the position of the columns to `"dodge"`, and be sure to add a label for the `fill` aesthetic. Adjust the caption to summarize your findings.

```{r}
#| eval: false
hmda_denial_by_race_sex |>
  ggplot(aes(x = derived_race, y= percent_denied, fill= derived_sex)) +
  geom_col(position= "dodge")+
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Percent of Individuals Denied In Race",
       fill= "Sex of Individuals",
       caption = str_wrap("This column plot visualizes the denial rates of home mortgage applications and sex of the applicants wo where rejected in MS in 2024. The plot shows that the highest number of join sex individuals were denied from the American Indian race and highest number of males and females were denied from 2 or more minority race", 120)) +
  theme_minimal() +
  coord_flip()
  
```

:::

::: {.callout-important}
###### Question 10

Create a plot below that visualizes how many applicants *of each race* were denied mortgages *for each denial reason*. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. You may wish to borrow some of your previous code to adjust the plot for legibility. I've given you some code below to filter the data frame to just those mortgage applications that were denied.

```{r}
#| eval: false

hmda_ms_2024 |>
  filter(action_taken == "Application denied") |>
 ggplot(aes(x = derived_race, fill= denial_reason)) +
  geom_bar(position= "dodge")+
  labs(title = "Number of Applicants per race who were denied for each reason",
       x = "Race of Applicant",
       y = "Count of Race Individuals",
       fill= "Reason for Loan Denial",
       caption = str_wrap("This barplot shows the reason why applicants of a particular race were denied the loans. It shows that the reason for most denials was multiple second by credit history which was most common in Black or African Americans",150))+
  coord_flip()+
  theme_minimal()+
  theme(axis.text.x = element_blank(),
    legend.position = "bottom",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"))
  
```

:::

::: {.callout-important}
###### Question 11

Below I've given you some code to produce a plot that visualizes the percent of mortgage applications that resulted in each action (e.g. loan originated, denied, etc.), broken out by the debt-to-income ratio of the applicant. Referring to last week's lab, add a line to this code to facet this plot by race. For which races was the denial rate higher for applicants with a lower than 50% debt-to-income ratio? Adjust the fact and interpretation presented in the caption with your response.

```{r}
#| eval: true

hmda_ms_2024 |>
  ggplot(aes(x = debt_to_income_ratio, fill = action_taken)) +
  geom_bar(position = "fill") +
  coord_flip() +
  #Add line to facet here
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Debt-to-Income Ratio",
       y = "Percent of Mortgage Applications",
       fill = "Action Taken", 
       caption = str_wrap("This bar plot visualizes the percent of mortgage applications that resulted in each action, broken out by the debt-to-income ratio of the applicant. From this plot, we can see that most applicants with a debt-to-income ratio of 50% or lesser were denied in the races American Indian, Black or African American and 2 or more minority groups.",width = 120)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        plot.caption = element_text(size = 8)) +
  guides(fill=guide_legend(nrow = 3))+
  facet_wrap(vars(derived_race))
```

:::


::: {.callout-important}
###### Question 12

Below I've given you some code to produce a plot that visualizes how the distributions of interest rates differed across debt-to-income ratios when loans were approved. Referring to last week's lab, add a line to this code to facet this plot by race. For which races were the interest rates higher even with lower debt-to-income ratios?  Adjust the fact and interpretation presented in the caption with your response.


```{r}
#| eval: true

hmda_ms_2024 |>
  filter(action_taken == "Loan originated") |>
  ggplot(aes(x = debt_to_income_ratio, y = as.numeric(interest_rate))) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Debt-to-Income Ratio",
       y = "Interest Rate",
       caption = str_wrap("These boxplots visualize the distribution of interest rates for loans originated in MS in 2024, broken out by the debt-to-income ratio of the applicant. From this plot, we can see that the median interest rate for borrowers with a debt-to-income ratio of under 50% was around 7%. Median interest rates increased for borrowers with a debt-to-income ratio of over 50%. This shows that a high debt-to-income ratio may be one factor contributing to higher interest rates. But Black or African Americans has an interest rate greater than 7% for a debt to income ration of under 50% which show some signs of racism.",width = 120)) +
  theme_minimal()+
  facet_wrap(vars(derived_race))
  
```

:::

::: callout-caution
## Ethical Considerations

This dataset makes loan-level data about individual sensitive financial information available to the public. To reduce the risk that borrowers may be identified in the data, the CFPB removes certain variables before publication, such as the loan's unique identifier, the property address, the applicant's credit score, and any free-form text that the applicant writes on the application. Applicants do not have the option to opt-out of their data being disclosed in this dataset. If it were possible to re-identify individual applicants in this dataset, what would be the risks to the public disclosure of their information? Which applicants would be the most vulnerable? How should we think about the balance between the social value of this public data disclosure and the possible risks to those represented in it?  Share your ideas on our `discussions` Slack channel.

#Indeed this dataset has some sensitive financial information like property value, det to income ratio, loan purpose and loan to value ratio. If it were possible to re-identify individual applicants, it would make individuals who took loans with a low debt to income ratio known to the public which is personal financial information that should not be known. It will make individuals loose trust in banks and make people shy away from loans when in deep financial crisis because of fear of their problem being made public. The most vulnerable applicants would be black or African Americans because they were recorded to have the highest denials due to credit history.I really like that this type of data is shared because it creates awareness about considerations and injustice when loans are granted and I also like the precautions taken to remove indentifiable variables which doesnt make any individual known. Even though the data discloses state location of the individuals which may create assumptions about the black people in the area having low credit history but no identifier was mentioned so I think the data has more positive social impact of creating awareness and reducing such injustice.
:::
